---
title: "Data Munging, Visualization, and Feature Engineering"
author: "Scott McKean"
date: "7/22/2020"
output: html_document
---

This notebook provides an introduction to data munging in the tidyverse. Why
tidyverse you might ask? Personally, I find it a really nice way to learn the 
'grammar of graphics' and 'grammar of data' - probably the best intro in my 
opinion, which is why I love teaching it.

The agenda for this 1 hour bootcamp includes:

1. Loading data
2. Tidy data concepts
3. Working with dates
4. Static vs. dynamic typed languages
4. Functional programming vs. object oriented programming
5. Data Structures

```{r}
library(tidyverse)
```

## Loading data

```{r}
treatments = readr::read_csv("./data/PerfTreatment_subset.csv")

well_header = readr::read_csv("./data/WellHeader_subset.csv")

production = readr::read_csv("./data/WellProduction_subset.csv")

```

## Tidy data concepts

`production` is a long table
`well_header` and `treatments` are wide tables

```{r}

```

## Working with dates

