<<<<<<< HEAD
---
title: "R Foundations"
author: "Scott McKean"
date: "7/15/2020"
output: html_document
---

This notebook provides some of the essentials to follow along with the 
first R bootcamp: R Fundamentals. The agenda for this 1 hour bootcamp includes:

1. R vs. Rstudio
2. Installing Packages (CRAN, versions, etc.)
3. Base R vs. Tidyverse vs. data.table
4. Static vs. dynamic typed languages
4. Functional programming vs. object oriented programming
5. Data Structures

There will actually be very little code in this bootcamp, with a focus more on
discussing high level concepts and providing tips to get going in the R universe.

## Installing R & RStudio

First, let's differentiate R and R Studio. 

R is an open-source statistical programming language. You can download it from:
https://www.r-project.org/

RStudio is a company that makes a great interactive development environment (IDE).
You can download the IDE here:
https://rstudio.com/products/rstudio/

I'll present a quick 5-10 min intro to the IDE and language. 

## Packages

Packages (sometimes called libraries) are the lifeblood of open-source programming.

A couple tips on packages:
- They are linked to specific versions of R
- They require periodic updates to stay 'current'
- You may want to install a specific version to avoid breaking changes
- Breaking changes are real, and the most annoying part of open-source

So where do you look for packages?
1. LMGTFY (let me google that for you)
2. https://cran.r-project.org/
3. Read vignettes - stupid word but really useful
https://cran.r-project.org/web/packages/dplyr/vignettes/dplyr.html

```{r}
# install a package
install.packages('tidyverse')
install.packages('data.table')

# load a package
library(tidyverse)
library(data.table)

# get some help
help(tidyverse)
help(dplyr)
```

## Base R vs. Tidyverse vs. data.table
R has a complicated history and has some 'camps' that don't always agree on how
things should be done. There are three main tribes:
1. Base R
2. Tidyverse
3. data.table

But, realistically, these tribes get along and play nicely, so don't be scared
to cross over. I'm going to explain a little of each using a builtin dataset.
As you can see - the approaches are really similar.

I'm a tidyverse 'fanatic' as it goes, so the next two sessions will generally use
tidyverse (and lots of pipes - love the pipe).

```{r}
cars_df = cars
summary(cars_df)
head(cars_df)

# base R
cars[cars$speed > 4,'dist']

# tidyverse
cars %>%
  dplyr::filter(speed > 4) %>%
  pull(dist)

# data.table
cars_dt = data.table::as.data.table(cars)
cars_dt[speed >= 4, list(dist)]$dist
```

## Static vs. dynamic typed languages
It's important to realize that R and Python are dynamically typed languages - 
which means that you don't find problems until you run a code chunk. It also
makes them much much nicer to iterate and code in (you'll understand if you 
ever did C++ or likewise in undergrad).

Also, it is best to assume that R makes a copy of almost everything, unlike 
Python and Java. So keep that in mind when you are doing referencing. Here is
a simple example.

```{r}
a = 2
a
a = 3
a
b = a
a = 5
c(a,b)
```

## Functional programming vs. object oriented programming
This is a little philosophical, but I think very important when you are starting
to program. There are two main ways to program - object oriented and functional.
You'll run across both, but R is a combination of both approaches. I prefer to
teach people functional programming first - it's less abstract and I find that it
is easier to understand for scientists and engineers.

Here is an example of a functional workflow with two custom functions
in the Tidyverse style.

```{r}
feature_eng <- function(df){
  return(df$speed*df$dist)
}

cars_feat <- cars %>%
  dplyr::mutate(speeddist = feature_eng(.))

adhoc_clust <- function(df){
  case_when(
    (df$speed>10)&(df$dist<20)   ~ 1,
    (df$speed>10)&(df$dist>20)   ~ 2,
    TRUE                         ~ 3
    )
}

super_pipe <- cars %>%
  dplyr::filter(speed > 10) %>%
  dplyr::mutate(speeddist = feature_eng(.)) %>%
  dplyr::mutate(adhoc_clust = adhoc_clust(.))
```

## Data Structures
My last bit of advice before questions has to do around data structures.
The heart of programming is data structures, which is why every introductory
course starts with them. Here are a couple tips:

1. Pass the simplest data structure possible to the function (an array instead of a dataframe).
2. Bundle your variables in higher level data structures (lists/dataframes)
3. Use consistent data structures throughout your analysis (dataframe vs. datatable)
4. Pay attention to naming!!! Names are important
5. I prefer the following structures in R, in order of preference 1) dataframe, 2) array, 3) list.

```{r}
# list
cars_list <- list(cars$speed, cars$dist, 'cars')
cars_list[1]

# array
array <- c(1,2,3)
array[2]

# dataframe
new_df = data.frame(x = seq(1,5), y=2, z='col')
```
=======
---
title: "R Foundations"
author: "Scott McKean"
date: "7/15/2020"
output: html_document
---

This notebook provides some of the essentials to follow along with the 
first R bootcamp: R Fundamentals. The agenda for this 1 hour bootcamp includes:

1. R vs. Rstudio
2. Installing Packages (CRAN, versions, etc.)
3. Base R vs. Tidyverse vs. data.table
4. Static vs. dynamic typed languages
4. Functional programming vs. object oriented programming
5. Data Structures

There will actually be very little code in this bootcamp, with a focus more on
discussing high level concepts and providing tips to get going in the R universe.

## Installing R & RStudio

First, let's differentiate R and R Studio. 

R is an open-source statistical programming language. You can download it from:
https://www.r-project.org/

RStudio is a company that makes a great interactive development environment (IDE).
You can download the IDE here:
https://rstudio.com/products/rstudio/

I'll present a quick 5-10 min intro to the IDE and language. 

## Packages

Packages (sometimes called libraries) are the lifeblood of open-source programming.

A couple tips on packages:
- They are linked to specific versions of R
- They require periodic updates to stay 'current'
- You may want to install a specific version to avoid breaking changes
- Breaking changes are real, and the most annoying part of open-source

So where do you look for packages?
1. LMGTFY (let me google that for you)
2. https://cran.r-project.org/
3. Read vignettes - stupid word but really useful
https://cran.r-project.org/web/packages/dplyr/vignettes/dplyr.html

```{r}
# install a package
install.packages('tidyverse')
install.packages('data.table')

# load a package
library(tidyverse)
library(data.table)

# get some help
help(tidyverse)
help(dplyr)
```

## Base R vs. Tidyverse vs. data.table
R has a complicated history and has some 'camps' that don't always agree on how
things should be done. There are three main tribes:
1. Base R
2. Tidyverse
3. data.table

But, realistically, these tribes get along and play nicely, so don't be scared
to cross over. I'm going to explain a little of each using a builtin dataset.
As you can see - the approaches are really similar.

I'm a tidyverse 'fanatic' as it goes, so the next two sessions will generally use
tidyverse (and lots of pipes - love the pipe).

```{r}
cars_df = cars
summary(cars_df)
head(cars_df)

# base R
cars[cars$speed > 4,'dist']

# tidyverse
cars %>%
  dplyr::filter(speed > 4) %>%
  pull(dist)

# data.table
cars_dt = data.table::as.data.table(cars)
cars_dt[speed >= 4, list(dist)]$dist
```

## Static vs. dynamic typed languages
It's important to realize that R and Python are dynamically typed languages - 
which means that you don't find problems until you run a code chunk. It also
makes them much much nicer to iterate and code in (you'll understand if you 
ever did C++ or likewise in undergrad).

Also, it is best to assume that R makes a copy of almost everything, unlike 
Python and Java. So keep that in mind when you are doing referencing. Here is
a simple example.

```{r}
a = 2
a
a = 3
a
b = a
a = 5
c(a,b)
```

## Functional programming vs. object oriented programming
This is a little philosophical, but I think very important when you are starting
to program. There are two main ways to program - object oriented and functional.
You'll run across both, but R is a combination of both approaches. I prefer to
teach people functional programming first - it's less abstract and I find that it
is easier to understand for scientists and engineers.

Here is an example of a functional workflow with two custom functions
in the Tidyverse style.

```{r}
feature_eng <- function(df){
  return(df$speed*df$dist)
}

cars_feat <- cars %>%
  dplyr::mutate(speeddist = feature_eng(.))

adhoc_clust <- function(df){
  case_when(
    (df$speed>10)&(df$dist<20)   ~ 1,
    (df$speed>10)&(df$dist>20)   ~ 2,
    TRUE                         ~ 3
    )
}

super_pipe <- cars %>%
  dplyr::filter(speed > 10) %>%
  dplyr::mutate(speeddist = feature_eng(.)) %>%
  dplyr::mutate(adhoc_clust = adhoc_clust(.))
```

## Data Structures
My last bit of advice before questions has to do around data structures.
The heart of programming is data structures, which is why every introductory
course starts with them. Here are a couple tips:

1. Pass the simplest data structure possible to the function (an array instead of a dataframe).
2. Bundle your variables in higher level data structures (lists/dataframes)
3. Use consistent data structures throughout your analysis (dataframe vs. datatable)
4. Pay attention to naming!!! Names are important
5. I prefer the following structures in R, in order of preference 1) dataframe, 2) array, 3) list.

```{r}
# list
cars_list <- list(cars$speed, cars$dist, 'cars')
cars_list[1]

# array
array <- c(1,2,3)
array[2]

# dataframe
new_df = data.frame(x = seq(1,5), y=2, z='col')
```
>>>>>>> 37c4018eed1a082e0081ceabc64c580a7884cc31
